---
title: "Assessment 2 GY476"
output:
  html_document: default
  pdf_document: default
---

# SETUP: Load packages and set directories

```{r setup}
# Set working directory
wdir <- "/Users/dylanhillier/Library/CloudStorage/OneDrive-LondonSchoolofEconomics/GY476_Geographic_Information_System/assessment02"

# Load required packages
if(!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)
library(ggplot2)
library(dplyr)
library(sf)
library(tmap)

# Create data directory if it doesn't exist
if(!dir.exists(file.path(wdir, "data"))) {
  dir.create(file.path(wdir, "data"))
}
ddir <- file.path(wdir, "data")
```

# DOWNLOAD DATA: Airbnb listings and neighbourhood boundaries

```{r download-data}
# Download Airbnb listings
download.file(
  url = "https://data.insideairbnb.com/united-states/ca/los-angeles/2025-09-01/visualisations/listings.csv",
  destfile = file.path(ddir, "listing.csv")
)

# Download neighbourhood boundaries
download.file(
  url = "https://data.insideairbnb.com/united-states/ca/los-angeles/2025-09-01/visualisations/neighbourhoods.geojson",
  destfile = file.path(ddir, "neighbourhoods.geojson")
)
```

# LOAD DATA: Read in all datasets

```{r load-data}
# Airbnb listings
airbnb_raw <- read.csv(file.path(ddir, "listing.csv"))

# Neighbourhood boundaries
neighbourhoods_sf <- st_read(file.path(ddir, "neighbourhoods.geojson"))

# LA census tract geometry
census_tracts_sf <- st_read(file.path(ddir, "LA_geometry.gpkg")) %>%
  st_transform(32611)
  

# Census data
census_data <- read.csv(file.path(ddir, "ACS_2019_2023_LA_vars.csv")) %>%
  tibble() %>%
  mutate(
    GEOID = as.character(GEOID),
    GEOID = paste0("06", str_sub(GEOID, start = 2))
  )
```

# JOIN DATA: Combine census geometry with census data

```{r join-census}
census_joined <- census_tracts_sf %>%
  left_join(census_data, by = "GEOID")
```



# PREPARE AIRBNB DATA: Remove outliers and convert to spatial


```{r prepare-airbnb}
# Calculate IQR bounds for outlier removal
Q1 <- quantile(airbnb_raw$price, 0.25, na.rm = TRUE)
Q3 <- quantile(airbnb_raw$price, 0.75, na.rm = TRUE)
IQR_val <- Q3 - Q1

# Filter outliers and convert to sf object
airbnb_sf <- airbnb_raw %>%
  tibble() %>%
  filter(
    price >= (Q1 - 1.5 * IQR_val) & 
    price <= (Q3 + 1.5 * IQR_val)
  ) %>%
  st_as_sf(coords = c("longitude", "latitude")) %>%
  st_set_crs(4326) %>%      
  st_transform(32611) 
```



# FIGURE 1: Airbnb Listings by Room Type

```{r figure-1}



# First, let's find the actual extent of Catalina Island
# Check the coordinates of Catalina listings
catalina_points <- airbnb_sf %>%
  filter(st_coordinates(.)[,2] < 3720000)  # Points south of mainland

# Get the bounding box of Catalina points
st_bbox(catalina_points)

# Define bounding boxes based on actual data
# Mainland LA (excluding Catalina)
# More centred view
mainland_bbox <- st_bbox(c(xmin = 320000, ymin = 3710000, 
                           xmax = 445000, ymax = 3860000), 
                         crs = 32611)

# Catalina Island - adjusted coordinates
catalina_bbox <- st_bbox(c(xmin = 350000, ymin = 3685000, 
                           xmax = 385000, ymax = 3715000), 
                         crs = 32611)

# Split by room type
airbnb_entire <- airbnb_sf %>%
  filter(room_type == "Entire home/apt")

airbnb_other <- airbnb_sf %>%
  filter(room_type %in% c("Hotel room", "Private room", "Shared room"))

# ---- MAP 1: Entire Home/Apartment ----
tmap_mode("plot")

# Main map (mainland)
main_entire <- tm_shape(census_tracts_sf, bbox = mainland_bbox) +
  tm_borders(col = "grey70", lwd = 0.3) +
tm_shape(airbnb_entire) +
  tm_dots(fill = "darkblue",
          size = 0.05,
          fill_alpha = 0.5) +
tm_title("Entire Home/Apartment", fontface = "bold", size = 1.2) +
tm_scalebar(position = c("left", "bottom"), width = 15) +
tm_compass(type = "arrow", position = c("right", "top"), size = 2.4) +
tm_layout(frame = TRUE)

# Inset map (Catalina)
inset_entire <- tm_shape(census_tracts_sf, bbox = catalina_bbox) +
  tm_fill(fill = "grey95") +
  tm_borders(col = "grey70", lwd = 0.5) +
tm_shape(airbnb_entire) +
  tm_dots(fill = "darkblue",
          size = 0.2,
          fill_alpha = 0.7) +
tm_title("Catalina Island", size = 0.7, fontface = "bold") +
tm_layout(frame = TRUE,
          frame.lwd = 2,
          bg.color = "white")

# ---- MAP 2: Hotel/Private/Shared Room ----

# Main map (mainland)
main_other <- tm_shape(census_tracts_sf, bbox = mainland_bbox) +
  tm_borders(col = "grey70", lwd = 0.3) +
tm_shape(airbnb_other) +
  tm_dots(fill = "darkred",
          size = 0.05,
          fill_alpha = 0.5) +
tm_title("Hotel / Private / Shared Room", fontface = "bold", size = 1.2) +
tm_scalebar(position = c("left", "bottom"), width = 15) +
tm_compass(type = "arrow", position = c("right", "top"), size = 2.4) +
tm_layout(frame = TRUE)

# Inset map (Catalina)
inset_other <- tm_shape(census_tracts_sf, bbox = catalina_bbox) +
  tm_fill(fill = "grey95") +
  tm_borders(col = "grey70", lwd = 0.5) +
tm_shape(airbnb_other) +
  tm_dots(fill = "darkred",
          size = 0.2,
          fill_alpha = 0.7) +
tm_title("Catalina Island", size = 0.7, fontface = "bold") +
tm_layout(frame = TRUE,
          frame.lwd = 2,
          bg.color = "white")


# ---- Save with grid viewports ----
library(grid)

png(file.path(ddir, "figure_1_room_types.png"), 
    width = 14, height = 8, units = "in", res = 300)

grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))

# Left panel
pushViewport(viewport(layout.pos.col = 1))
print(main_entire, vp = viewport(x = 0.5, y = 0.5, width = 0.95, height = 0.95))
print(inset_entire, vp = viewport(x = 0.78, y = 0.17, width = 0.22, height = 0.28))
popViewport()

# Right panel
pushViewport(viewport(layout.pos.col = 2))
print(main_other, vp = viewport(x = 0.5, y = 0.5, width = 0.95, height = 0.95))
print(inset_other, vp = viewport(x = 0.78, y = 0.17, width = 0.22, height = 0.28))
popViewport()

dev.off()


```

# MAP 2: Static map of listings with neighbourhood boundaries

```{r map-listings}
tmap_mode("plot")

map_listings <- tm_shape(neighbourhoods_sf) +
  tm_polygons(
    fill = "#f7f7f7",
    col = "#969696",
    lwd = 0.5,
    fill_alpha = 0.8
  ) +
tm_shape(airbnb_sf) +
  tm_dots(
    fill = "price",
    fill.scale = tm_scale_continuous(values = "YlOrRd"),
    fill.legend = tm_legend(title = "Price ($)", frame = FALSE),
    size = 0.15,
    col = "#00000020"
  ) +
tm_title("Airbnb Listing Prices in Los Angeles") +
tm_compass(type = "arrow", position = c("right", "top"), size = 1.5) +
tm_scalebar(position = c("left", "bottom"), width = 15) +
tm_layout(
  frame = FALSE,
  legend.outside = TRUE,
  legend.outside.position = "right"
)

map_listings

# Save map
tmap_save(
  map_listings,
  filename = file.path(ddir, "la_airbnb_map.png"),
  width = 10, height = 8, units = "in", dpi = 300, device = png
)
```

# MAP 3: Choropleth of mean price by neighbourhood

```{r map-choropleth}
# Fix invalid geometries
census_tracts_sf <- st_make_valid(census_tracts_sf)
airbnb_sf <- st_make_valid(airbnb_sf)

# Calculate mean price per census tract
tract_prices <- airbnb_sf %>%
  st_join(census_tracts_sf) %>%
  st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarise(mean_price = mean(price, na.rm = TRUE))

# Join prices back to census tract geometry
tracts_with_prices <- census_tracts_sf %>%
  left_join(tract_prices, by = "GEOID")

# Identify top 3 and bottom 3 tracts by price
all_highlight <- tracts_with_prices %>%
  filter(!is.na(mean_price)) %>%
  mutate(price_rank = rank(-mean_price, ties.method = "first")) %>%
  filter(price_rank <= 3 | price_rank >= (max(price_rank) - 2)) %>%
  mutate(
    type = ifelse(price_rank <= 3, "Highest", "Lowest"),
    label = case_when(
      price_rank == 1 ~ "1st Highest",
      price_rank == 2 ~ "2nd Highest",
      price_rank == 3 ~ "3rd Highest",
      price_rank == max(price_rank) ~ "1st Lowest",
      price_rank == max(price_rank) - 1 ~ "2nd Lowest",
      price_rank == max(price_rank) - 2 ~ "3rd Lowest"
    )
  ) %>%
  st_centroid()

# Split into highest and lowest
highest_tracts <- all_highlight %>% 
  filter(type == "Highest") %>%
  mutate(label = factor(label, levels = c("1st Highest", "2nd Highest", "3rd Highest")))

lowest_tracts <- all_highlight %>% 
  filter(type == "Lowest") %>%
  mutate(label = factor(label, levels = c("1st Lowest", "2nd Lowest", "3rd Lowest")))

# Define bounding boxes
mainland_bbox <- st_bbox(c(xmin = 320000, ymin = 3710000, 
                           xmax = 445000, ymax = 3860000), 
                         crs = 32611)

catalina_bbox <- st_bbox(c(xmin = 350000, ymin = 3685000, 
                           xmax = 385000, ymax = 3715000), 
                         crs = 32611)

# ---- Main Map (Mainland) ----
tmap_mode("plot")

main_choropleth <- tm_shape(tracts_with_prices, bbox = mainland_bbox) +
  tm_fill(
    fill = "mean_price",
    fill.scale = tm_scale_continuous(values = "YlOrRd"),
    fill.legend = tm_legend(title = "Mean Price ($)")
  ) +
  tm_borders(col = "grey70", lwd = 0.2) +
  # Highest price tracts - GOLD
  tm_shape(highest_tracts) +
  tm_symbols(
    shape = "label",
    shape.scale = tm_scale_categorical(values = c(24, 25, 23)),
    shape.legend = tm_legend(title = "Highest Price"),
    fill = "#DAA520",
    size = 0.8,
    col = "black",
    lwd = 1.2
  ) +
  # Lowest price tracts - RED
  tm_shape(lowest_tracts) +
  tm_symbols(
    shape = "label",
    shape.scale = tm_scale_categorical(values = c(21, 22, 8)),
    shape.legend = tm_legend(title = "Lowest Price"),
    fill = "#ee1616",
    size = 0.8,
    col = "black",
    lwd = 1.2
  ) +
  tm_title("Mean Airbnb Price by Census Tract", fontface = "bold", size = 1.2) +
  tm_compass(type = "arrow", position = c("right", "top"), size = 2.4) +
  tm_scalebar(position = c("left", "bottom"), width = 15) +
  tm_layout(
    frame = TRUE,
    legend.outside = TRUE,
    legend.outside.position = "right"
  )

# ---- Inset Map (Catalina Island) ----
inset_choropleth <- tm_shape(tracts_with_prices, bbox = catalina_bbox) +
  tm_fill(
    fill = "mean_price",
    fill.scale = tm_scale_continuous(values = "YlOrRd"),
    fill.legend = NULL
  ) +
  tm_borders(col = "grey70", lwd = 0.3) +
  tm_title("Catalina Island", size = 1.0, fontface = "bold") +
  tm_layout(
    frame = TRUE,
    frame.lwd = 2,
    bg.color = "white",
    legend.show = FALSE
  )

# ---- Save with grid viewports ----
library(grid)

png(file.path(ddir, "figure_2_choropleth.png"), 
    width = 12, height = 10, units = "in", res = 300)

grid.newpage()
print(main_choropleth, vp = viewport(x = 0.5, y = 0.5, width = 1, height = 1))
print(inset_choropleth, vp = viewport(x = 0.87, y = 0.145, width = 0.25, height = 0.25))

dev.off()

main_choropleth
```

# Map 4 Census data

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidycensus)
library(tidyverse)
library(sf)
library(viridis)

census_api_key("c990ed72472399dd626592b06afee9a188d03320", install = TRUE, overwrite = TRUE)
options(tigris_use_cache = TRUE)

view_vars <- load_variables(2023, "acs5", cache = TRUE)

renter_data <- get_acs(
  state = "CA",
  county = "Los Angeles",
  geography = "tract",
  variables = c(total_units = "B25003_001", renter_occupied = "B25003_003"),
  geometry = TRUE,
  year = 2023,
  output = "wide"
) %>% tibble()
```

```{r}
renter_data <- renter_data %>%
  mutate(pct_renter = (renter_occupiedE / total_unitsE) * 100) %>%
  select(GEOID, pct_renter)

renter_sf <- census_tracts_sf %>%
  left_join(renter_data, by = "GEOID") %>%
  st_make_valid() %>%
  st_transform(crs = 32611)  # CRS transformation here

tmap_mode = "plot"
mainland_bbox <- st_bbox(c(xmin = 320000, ymin = 3710000, 
                           xmax = 445000, ymax = 3860000), 
                         crs = 32611)


catalina_bbox <- st_bbox(c(xmin = 340000, ymin = 3678000, 
                           xmax = 390000, ymax = 3708000), 
                         crs = 32611)

# Main map (mainland)
main_renter <- tm_shape(renter_sf, bbox = mainland_bbox) +
  tm_polygons(
    fill = "pct_renter",
    fill.scale = tm_scale_continuous(values = "YlOrRd"),
    fill.legend = tm_legend(title = "Percentage Renters (%)"),
    col = "white",
    lwd = 0.3
  ) +
tm_title("Percentage of Renters by Census Tract", fontface = "bold", size = 1.2) +
tm_compass(type = "arrow", position = c("right", "top"), size = 1.5) +
tm_scalebar(position = c("left", "bottom"), width = 15) +
tm_layout(frame = TRUE)

# Inset map (Catalina)
inset_renter <- tm_shape(renter_sf, bbox = catalina_bbox) +
  tm_polygons(
    fill = "pct_renter",
    fill.scale = tm_scale_continuous(values = "YlOrRd"),
    fill.legend = tm_legend(show = FALSE),
    col = "white",
    lwd = 0.3
  ) +
tm_title("Catalina Island", size = 0.7, fontface = "bold") +
tm_layout(frame = TRUE,
          frame.lwd = 2,
          bg.color = "white")

# Save with grid viewports
library(grid)

png(file.path(ddir, "renter_map.png"), 
    width = 10, height = 8, units = "in", res = 300)

grid.newpage()
print(main_renter, vp = viewport(x = 0.5, y = 0.5, width = 1, height = 1))
print(inset_renter, vp = viewport(x = 0.815, y = 0.1325, width = 0.22, height = 0.28))

dev.off()
```